---
title: "HW5"
# subtitle: "possible subtitle goes here"
author:
  - Cheng Huang 2658312
    
date: "`r format(Sys.time(), '%d %B %Y')`"
documentclass: article
papersize: letter
fontsize: 11pt
bibliography: template.bib
biblio-style: asa
output:
  pdf_document: default
  html_document:
    df_print: paged
always_allow_html: yes
---

#  verify the validity of the provided E- and M-steps
\[Q(\Psi|\Psi^{(k)})=\sum_zp(z|x,y, \Psi^{(k)})ln p(x,y,z|\Psi)\]
\[=\sum_{i=1}^n\sum_{j=1}^mp(z_{ij}|(x,y, \Psi^{(k)}))z_{ij}\{log\pi_i+log\varphi(y_i-x_i^T\beta_j;0, \sigma^2)\}\]
\[=\sum_{i=1}^n\sum_{j=1}^m E(z_{ij}|x,y, \Psi^{(k)})\{log\pi_i+log\varphi(y_i-x_i^T\beta_j;0, \sigma^2)\}\]

Since $z_{ij}$ is 1 if the ith observatiojn is from the jth group, 0 otherwise. So
\[E(z_{ij}|x, y, \Psi^{(k)}) = \sum_z z_{ij}p(z_{ij}|(x,y, \Psi^{(k)})) = p(z_{ij}=1|x,y, \Psi^{(k)}) \]
\[= \frac{\pi_j^{(k)} \varphi(y_i-x_i^T\beta_j^{(k)}; 0, \sigma^{2(k)})}{\sum_{j=1}^{m}\pi_j^{(k)} \varphi(y_i-x_i^T\beta_j^{(k)}; 0, \sigma^{2(k)})}\]

Maximize $Q(\Psi|\Psi^{(k)})$ respect to $\pi_j^{(k+1)}, \beta_j^{(k+1)}, \sigma^{2(k+1)}$:\
$\pi_j^{(k+1)}$:
\[\frac{\partial Q(\Psi|\Psi^{(k)})}{\partial \pi_j} = \frac{ \sum_{i=1}^n\sum_{j=1}^mp_{ij}^{(k+1)}}{\pi_j} = 0\]
So $\pi_j^{(k+1)} = \frac{\sum_{i=1}^np_{ij}^{(k+1)}}{n}$.\
$\beta_j^{(k+1)}$:\
\[log \varphi(y_i-x_i^T\beta_j;0|\sigma^2)=-\frac{1}{2}log(2\pi\sigma^2)-\frac{(y_i-x_i^T\beta_j)^2}{2\sigma^2}\]
substitute above equation to $Q(\Psi|\Psi^{(k)})$.

\[Q(\Psi|\Psi^{(k)})=\sum_{i=1}^n\sum_{j=1}^mp_{ij}^{(k+1)}\{log \pi_j - \frac{1}{2}log(2\pi\sigma^2)-\frac{(y_i - x_i^T\beta_j)^2}{2\sigma^2}\}\]
\[\frac{\partial Q(\Psi|\Psi^{(k)})}{\partial \beta_j} = 0\]
gives that $\beta_j^{(k+1)} = (\sum_{i=1}^n x_i x_i^T p_{ij}^{(k+1)} )^{-1} (\sum_{i=1}^n x_i p_{ij}^{(k+1)} y_i)$.
\[\frac{\partial Q(\Psi|\Psi^{(k)})}{\partial \sigma^2} = 0\]
gives that $\sigma^{2^{(k+1)}} = \frac{\sum_{i=1}^n \sum_{j=1}^m p_{ij}^{(k+1)} (y_i - \mathbf{x_i}^T \beta_j^{(k+1)})^2}{n}$.


# Implement algorithm

```{r em, echo = TRUE, message = FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE) 
require(data.table)
require(knitr)
require(pracma) #NR
require(ggplot2)

regmix_em <- function(y, xmat, pi.init, beta.init, sigma.init, control = list(maxiter = 100, tol = .Machine$double.eps ^ 0.5)){
  tol <- control$tol
  maxiter <- control$maxiter
  
  xmat <- as.matrix(xmat)
  n <- nrow(xmat)
  m <- length(pi.init)
  p <- ncol(xmat)
  pi <- pi.init
  beta <- beta.init
  sigma <- sigma.init
  pij <- matrix(0, nrow = n, ncol = m)
  pi.new <- rep(0, m)
  beta.new <- matrix(0, nrow = p, ncol = m)
  converge <- 1
  
  iter <- 1  

  while ((converge > tol) && (iter < maxiter)){
    
    for (i in 1:n){
      ##print(length(xmat[i, ]))
      ##print(length(beta))
     pij[i, ] <- (pi * dnorm (y[i] - xmat[i, ] %*% beta, mean = 0, sd = sigma))/(sum(pi * dnorm (y[i] - xmat[i, ] %*% beta, mean = 0, sd = sigma)))
    }
    pi.new <- colMeans(pij)
    for (j in 1:m){
      beta.new[, j] <- solve(t(xmat) %*% diag (pij[, j])%*% xmat) %*% t(xmat) %*% diag (pij[ ,j]) %*% y
    }
    
    sigma.new <- sqrt(sum(pij * (y %*% t(rep(1, m)) - xmat %*% beta.new) ^ 2)/n)
    
    converge <- sum(abs(pi.new - pi)) + sum(abs(beta.new - beta))+ abs (sigma.new -sigma)
    
    iter <- iter + 1
    #if(iter >= maxiter) break
    pi <- pi.new
    beta <- beta.new
    sigma <- sigma.new
  }
  return(list(pi = pi.new, beta = beta.new, sigma = sigma.new, convergence = converge))
}

```
# generate data
```{r em2, echo = TRUE, message = FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE) 
regmix_sim <- function(n, pi, beta, sigma) {
    K <- ncol(beta)
    p <- NROW(beta)
    xmat <- matrix(rnorm(n * p), n, p) # normal covaraites
    error <- matrix(rnorm(n * K, sd = sigma), n, K)
    ymat <- xmat %*% beta + error # n by K matrix
    ind <- t(rmultinom(n, size = 1, prob = pi))
    y <- rowSums(ymat * ind)
    data.frame(y, xmat)
}
n <- 400
pi <- c(.3, .4, .3)
bet <- matrix(c( 1,  1,  1, 
                -1, -1, -1), 2, 3)
sig <- 1
set.seed(1205)
dat <- regmix_sim(n, pi, bet, sig)

regmix_em(y = dat[,1], xmat = dat[,-1], 
          pi.init = pi / pi / length(pi),
          beta.init = matrix(c( 1,  2,  3, 
                -1, -1, -1), 2, 3),
          sigma.init = sig / sig, 
          control = list(maxiter = 500, tol = 1e-5))
```


# Reference {-}

https://stackoverflow.com/questions/19079152/contour-plot-of-a-custom-function-in-r
[jun-yan/stat-5361]https://github.com/jun-yan/stat-5361 \

